Let's consider the crabs data set we studied in lab. Recall that species(sp) is a factor with 2 levels Blue(B) and Orange(O). We want to study the difference of frontal lobe size(FL) of two species. 

```{r}
library(MASS)
data(crabs)
```
```{r}
head(crabs)
```

Consider the model $Y_i = \mu_1x_{Bi} + \mu_2x_{Oi} + \epsilon_i$,$i=1,...,200$. $Y_i$ is the FL of observation i. $x_{Bi}$ is 1 if $sp=B$ for obervation i and 0 otherwise. Similarly, $x_{Oi}$ is 1 if $sp=O$ for obervation i and 0 otherwise. $\epsilon_i$ are i.i.d with mean 0 and variance $\sigma^2$. This model can be fitted to the `crabs` dataset in R using the `lm()` function. The resulting summary is provided below.
```{r}
lm_crab <- lm(FL~sp-1, data=crabs)
summary(lm_crab)
```

(a). What do $\mu_1$ and $\mu_2$ measure in the above probability model?

(b). Build a 95% confidence interval for $\mu_1$ using normal approximation

(c). Recall in homework we know that the full estimated covariance matrix of $\mathbf{\hat{\mu}} = (\hat{\mu}_1,\hat{\mu}_2)$ can be found by  
```{r}
V <- summary(lm_crab)$cov.unscaled * summary(lm_crab)$s^2
V
```
Use V and information provided in summary(lm_crab) to write down an expression that constructs a 95% confidence interval for $\mu_1-\mu_2$.
