Recall the nels88 dataset from lab. These data are a subset of the National Education Longitudinal Study of 1988 which examined schoolchildren’s performance on a math test score in 8th grade. `ses` is the socioeconomic status of parents and `paredu` is the parents highest level of education achieved (less than high school, high school, college, BA, MA, PhD). The data were as follows:
```{r lm_nels88_example}
library(faraway)
data(nels88)
head(nels88)
```
We fit a regression model to the data. The rounded co-efficients for the model are provided below:
```{r}
fit <- lm(math ~ ses + paredu, data = nels88)
round(summary(fit)$coef)
```

(a)  Describe a suitable probability model, in matrix form, to give a sample version of the linear model that has been fit above.

(b) Find the predicted math score for a student whose family has an ses value of -0.5 and whose parents' highest education level is high school (`hs`).

(c) How do you calculate the residual standard error of this model? (Give the formula).

-------


