---
title: "Quiz 2, STATS 401 W18"
author: "In lab on 3/29 or 3/30"
output:
  pdf_document:
  toc: no
header-includes:
   - \usepackage{ifthen}
csl: ecology.csl
---

```{r preliminaries,echo=F,cache=F}
ALL <- TRUE
ALL <- FALSE
SOL <- FALSE
#SOL <- TRUE
set.seed(12)
library(mat2tex)
```

\ifthenelse{\isodd{0}}{
 \newcommand\exam[1]{#1} %% exam paper
 \newcommand\solution[1]{}
}{ 
 \newcommand\exam[1]{} %% solutions
 \newcommand\solution[1]{ {\bf Solution}. {#1}}
} 

\newcommand\mat[1]{\mathbb{#1}}
\newcommand\vect[1]{\mathbf{#1}}
\newcommand\transpose{{\scriptscriptstyle \mathrm{T}}}
\newcommand\prob{\mathrm{P}}
\newcommand\E{\mathrm{E}}
\newcommand\SE{\mathrm{SE}}
\newcommand\var{\mathrm{Var}}
\newcommand\cov{\mathrm{Cov}}



**This document generates different random quizzes each time it is compiled by Rmarkdown. The actual quiz will be a realization generated by this random process, or something very similar.**

**Instructions. You have a time allowance of 50 minutes, though the quiz may take you much less time and you can leave lab once you are done.  The quiz is closed book. Any electronic devices in your possession must be turned off and remain in a bag on the floor.**

---------------

### Definitions

D1. What is a p-value? What is a 95% confidence interval? 

------------

### Mean and variance calculations

S1. A basic exercise.
```{r s1-all-list,include=F}
s1_files <- list.files(path="s1",full.names=T,pattern="*.Rmd$") 
```
```{r s1, child = sample(s1_files,size=1),eval=!ALL}
```
```{r s1-1, child = s1_files[1],eval=ALL}
```
```{r s1-2, child = s1_files[2],eval=ALL}
```
```{r s1-3, child = s1_files[3],eval=ALL}
```
```{r s1-4, child = s1_files[4],eval=ALL}
```
```{r s1-5, child = s1_files[5],eval=ALL}
```
```{r s1-6, child = s1_files[6],eval=ALL}
```

S2. An example involving sums of squares and products.
```{r s2-all-list,include=F}
s2_files <- list.files(path="s2",full.names=T,pattern="*.Rmd$")
```
```{r s2, child = sample(s2_files,size=1),eval=!ALL}
```
```{r s2-1, child = s2_files[1],eval=ALL}
```
```{r s2-2, child = s2_files[2],eval=ALL}
```
```{r s2-3, child = s2_files[3],eval=ALL}
```
```{r s2-4, child = s2_files[4],eval=ALL}
```
```{r s2-5, child = s2_files[5],eval=ALL}
```
```{r s2-6, child = s2_files[6],eval=ALL}
```
```{r s2-7, child = s2_files[7],eval=ALL}
```


###  Confidence intervals for linear models

You are expected to be able to construct a confidence interval for a coefficient in a linear model, and to give a correct technical description of how the interval is contructed and what the interval means.

C1. Constructing a confidence interval for a parameter using R output.


C2. Explaining the construction and interpretation of the confidence interval.

Which of the following is a correct interpretation of a CI?

Which of the following is a correct construction of a CI for a linear model?

### Prediction for linear models

```{r f-all-list,include=F}
f_files <- list.files(path="f",full.names=T,pattern="*.Rmd$")
```
```{r f, child = sample(f_files,size=1),eval=!ALL}
```
```{r f-1, child = f_files[1],eval=ALL}
```
```{r f-2, child = f_files[2],eval=ALL}
```
```{r f-3, child = f_files[3],eval=ALL}
```

### Hypothesis tests for a single parameter

H1. Set up and make a hypothesis test for a single parameter, with help from R

H2. Set up and make a hypothesis test for a group of parameters.

### F tests for groups of parameters

F1. The following model was fitted to a dataset  Write out the null and alternative hypothesis, the test statistic, the p-value and your conclusion.

------



License: This material is provided under an [MIT license](https://ionides.github.io/401w18/LICENSE)<br>




------
