---
title: "Quiz 2, STATS 401 W18"
author: "In lab on 3/29 or 3/30"
output:
  pdf_document:
  toc: no
header-includes:
   - \usepackage{ifthen}
csl: ecology.csl
---

```{r preliminaries,echo=F,cache=F}
ALL <- TRUE
#ALL <- FALSE
SOL <- FALSE
#SOL <- TRUE
set.seed(12)
library(mat2tex)
```

\ifthenelse{\isodd{0}}{
 \newcommand\exam[1]{#1} %% exam paper
 \newcommand\solution[1]{}
}{ 
 \newcommand\exam[1]{} %% solutions
 \newcommand\solution[1]{ {\bf Solution}. {#1}}
} 

\newcommand\mat[1]{\mathbb{#1}}
\newcommand\vect[1]{\mathbf{#1}}
\newcommand\transpose{{\scriptscriptstyle \mathrm{T}}}
\newcommand\prob{\mathrm{P}}
\newcommand\E{\mathrm{E}}
\newcommand\SE{\mathrm{SE}}
\newcommand\var{\mathrm{Var}}
\newcommand\cov{\mathrm{Cov}}



**This document produces different random quizzes each time the source code generating it is run. The actual quiz will be a realization generated by this random process, or something similar.**

**Instructions. You have a time allowance of 50 minutes, though the quiz may take you much less time and you can leave lab once you are done.  The quiz is closed book, and you are not allowed access to any notes. Any electronic devices in your possession must be turned off and remain in a bag on the floor.**

### Formulas
  
* The following formulas will be provided. To use these formulas properly, you need to make appropriate definitions of the necessary quantities.

(1)$\quad\quad\mathbf{b}=\big(\mathbb{X}^\transpose \mathbb{X} \big)^{-1}\, \mathbb{X}^\transpose\mathbf{y}$

(2)$\quad\quad\var(\hat{\mathbf{\beta}})= \sigma^2 \big( \mathbb{X}^\transpose \mathbb{X} \big)^{-1}$

(3)$\quad\quad\var(\mathbb{A}\mathbf{Y})=\mathbb{A}\var(\mathbf{Y})\mathbb{A}^\top$

(4)$\quad\quad\var(X)=\E\big[ (X-\E[X])^2\big] = \E[X^2]-\big(\E[X]\big)^2$

(5)$\quad\quad\cov(X,Y)=\E\big[ \big(X-\E[X]\big)\big(Y-\E[Y]\big)\big] = \E[XY]-\E[X]\,\E[Y]$

(6)$\quad\quad$
The binomial ($n,p$) distribution has mean $np$ and variance $np(1-p)$.

\vspace{1cm}

From `?pnorm`:
```
pnorm(q, mean = 0, sd = 1)
qnorm(p, mean = 0, sd = 1)
q: vector of quantiles.
p: vector of probabilities.
```

---------------


### Q1. Calculating means and variances, and making a normal approximation

```{r q1-all-list,include=F}
q1_files <- list.files(path="q1",full.names=T,pattern="*.Rmd$") 
```
```{r q1, child = sample(q1_files,size=1),eval=!ALL}
```
```{r q1-1, child = q1_files[1],eval=ALL}
```
```{r q1-2, child = q1_files[2],eval=ALL}
```
```{r q1-3, child = q1_files[3],eval=ALL}
```


### Q2. Prediction using a linear model
```{r q2-all-list,include=F}
q2_files <- list.files(path="q2",full.names=T,pattern="*.Rmd$")
```
```{r q2, child = sample(q2_files,size=1),eval=!ALL}
```
```{r q2-1, child = q2_files[1],eval=ALL}
```
```{r q2-2, child = q2_files[2],eval=ALL}
```


### Q3. Comparing means using a linear model

This question will be based on HW7. It will involve either confidence interval construction or making a  hypothesis testing when comparing means of two samples in the context of a linear model.


```{r q3-all-list,include=F}
q3_files <- list.files(path="q3",full.names=T,pattern="*.Rmd$")
```
```{r q3, child = sample(q3_files,size=1),eval=!ALL}
```
```{r q3-1, child = q3_files[1],eval=ALL}
```
```{r q3-2, child = q3_files[2],eval=ALL}
```
```{r q3-3, child = q3_files[3],eval=ALL}
```

### Q4. Making and interpreting an F test


```{r q4-all-list,include=F}
q4_files <- list.files(path="q4",full.names=T,pattern="*.Rmd$")
```
```{r q4, child = sample(q4_files,size=1),eval=!ALL}
```
```{r q4-1, child = q4_files[1],eval=ALL}
```
```{r q4-2, child = q4_files[2],eval=ALL}
```

------



License: This material is provided under an [MIT license](https://ionides.github.io/401w18/LICENSE)<br>




------
