---
title: "Stats 401 Lab 2"
author: "401 GSI team"
date: "1/9/2018"
output:
  beamer_presentation:
    colortheme: dolphin
    incremental: no
  ioslides_presentation:
    incremental: no
  slidy_presentation:
    incremental: no
---
# GSI Office hour

At 2165 USB, the Science Learning Center Annex

- Yuan Sun. Mon 9-11 am
- Sanjana Gupta. Tue 9-11 am
- Naomi Giertych. Thu 9-11 am

# Homework

- Out of 10 points
- 1 point for the statement of sources
- 1 point for feedback
- Provide the code if the question request

# Swirl tutorial

We have finished lesson 1/3/4 in HW1.

- Any techical difficulties encountered working with swirl?

- Any questions about materials introduced in the tutorial?


# Swirl tutorial

You are asked to complete lesson 5/6/7/9 for HW2 and lesson 9 can be a little bit harder. 

- We can go through parts of it together at the end of this lab (if we have time).

- You can always go to our office hour for help.
 
# Basic matrix computation

- Addition
- Scalar multiplication
- Transpose
- Matrix multiplication
- Inverse
- Solving linear equations

# Addition

Let $\mathbb{A}=[a_{ij}]_{n\times p}$ and $\mathbb{B}=[b_{ij}]_{n\times p}$, then $\mathbb{A} + \mathbb{B}=[a_{ij} + b_{ij}]_{n\times p}$

For example, 
$$\mathbb{A}=
\begin{bmatrix}
    a_{11} &  a_{12}\\
    a_{21} &  a_{22}
\end{bmatrix}
$$
and 
$$\mathbb{B}=
\begin{bmatrix}
    b_{11} &  b_{12}\\
    b_{21} &  b_{22}
\end{bmatrix}
$$

Then
$$
\mathbb{A}+\mathbb{B}=
\begin{bmatrix}
    a_{11}+b_{11} &  a_{12}+b_{12}\\
    a_{21}+b_{21} &  a_{22}+b_{22}
\end{bmatrix}
$$

# Addition
```{r, eval = T}
# generate matrix A and B
A = matrix(c(3,-2,-1,4,1,2),nrow=2);A

B = matrix(1:6,nrow=2);B

A + B
```
# Scalar multiplication
Let $\mathbb{A}=[a_{ij}]_{n\times p}$, s be a scalar, then $s\mathbb{A}=[sa_{ij}]_{n\times p}$.

For example, 
$$\mathbb{A}=
\begin{bmatrix}
    a_{11} &  a_{12}\\
    a_{21} &  a_{22}
\end{bmatrix}
$$

Then
$$
s\mathbb{A}=
\begin{bmatrix}
    sa_{11} &  sa_{12}\\
    sa_{21} &  sa_{22}
\end{bmatrix}
$$

# Scalar multiplication
```{r, eval = T}
# Use same matrix A
A

# 5 time A
5 * A
```
# Transpose
Let $\mathbb{A}=[a_{ij}]_{n\times p}$, then $\mathbb{A}^{\top}=[a_{ji}]_{p\times n}$

For example, 
$$\mathbb{A}=
\begin{bmatrix}
    a_{11} &  a_{12}\\
    a_{21} &  a_{22}
\end{bmatrix}
$$


Then
$$\mathbb{A}^{\top}=
\begin{bmatrix}
    a_{11} &  a_{21}\\
    a_{12} &  a_{22}
\end{bmatrix}
$$

# Transpose
```{r, eval = T}
# Recall we have matrix A
A

# A transpose
C = t(A);C
```

# Matrix multiplication
Let $\mathbb{A}=[a_{ij}]_{n\times p}$ and $\mathbb{B}=[b_{ij}]_{p\times q}$, then $\mathbb{A}\mathbb{B}=[c_{ij}]_{n\times q}$ where $c_{ij} = \sum_{k=1}^p a_{ik}b_{kj}$

For example, 
$$\mathbb{A}=
\begin{bmatrix}
    a_{11} &  a_{12}\\
    a_{21} &  a_{22}
\end{bmatrix}
$$
and 
$$\mathbb{B}=
\begin{bmatrix}
    b_{11} &  b_{12}\\
    b_{21} &  b_{22}
\end{bmatrix}
$$

Then
$$
\mathbb{A}\mathbb{B}=
\begin{bmatrix}
    a_{11}b_{11}+a_{12}b_{21} &  a_{11}b_{12}+a_{12}b_{22}\\
    a_{21}b_{11}+a_{22}b_{21} &  a_{21}b_{12}+a_{22}b_{22}
\end{bmatrix}
$$

# Matrix multiplication
```{r, eval = T}
# Recall we have matrix B and C
B

C
# Let's caculate BC by hand
```
# Matrix multiplication
```{r, eval = T}
# Check with R
B %*% C

# notice that matrix multiplication is not commutative
C %*% B
```
# Inverse
$$\mathbb{A}=
\begin{bmatrix}
    a_{11} &  a_{12}\\
    a_{21} &  a_{22}
\end{bmatrix}
$$

Then
$$\mathbb{A}^{-1}=\frac{1}{a_{11}a_{22}-a_{12}a_{21}}
\begin{bmatrix}
    a_{22} &  -a_{12}\\
    -a_{21} &  a_{11}
\end{bmatrix}
$$

$det(\mathbb{A}) = a_{11}a_{22}-a_{12}a_{21}$ is called the determinant of $\mathbb{A}$. 

Need $det(\mathbb{A}) \neq 0$ for $\mathbb{A}$ to be invertible.

We only need to caculate the inverse a 2 by 2 matrix by hand 

# Inverse
```{r, eval = T}
# We can inverse higher dimensional matrix using R
# Produce a 3 by 3 matrix
set.seed(2018)
D = matrix(rnorm(9), nrow=3);D

# Inverse of D
solve(D)

```

# Solving linear equation
Suppose we want to solve $\mathbb{A}\mathbf{x} = \mathbf{b}$, then $\mathbf{x} = \mathbb{A}^{-1} \mathbf{b}$. (Assuming $\mathbb{A}$ is invertible)

As a example question, we want to solving the following linear equations in R
$$
\begin{array}{ccccccc}
x &+& y &&  &=& 2 \\
x &+& 2y &+& z &=& 1 \\
3x && &+& 2z &=& -3 \\
\end{array}$$

```{r, eval = T}
# This is the A we what
A = matrix(c(1,1,3,1,2,0,0,1,2),nrow=3);A

```

# Solving linear equations
```{r, eval = T}
# This is the b we what
b = c(2,1,-3)

# solve for x
x = solve(A) %*% b;x
```

# In lab activity
```{r, eval = FALSE}
# We generate the data similar as the homework
randomMatrix <- function(p,q,values=-4:4){
   matrix(sample(values,size=p*q,replace=TRUE),p,q)
}
set.seed(2018) 
A <- randomMatrix(2,2)
B <- randomMatrix(2,2)
```
1. Caculate the following by hand and check your results with R
- $\mathbb{A} + \mathbb{B}$
- $\mathbb{A}\mathbb{B}$
- $\mathbb{A}^{-1}$

2. Solve the following linear equations with R
$$
\begin{array}{ccccccc}
-x &+& y &+& z &=& 1.5 \\
x &+& 2y &-& z &=& -2 \\
3x && &+& 2z &=& -3 \\
\end{array}$$

# Lab ticket
1. Suppose $\mathbb{A}$ is a $4 \times 6$ matrix and $\mathbb{B}$ is a $3 \times 6$ matrix. What is the dimension of $\mathbb{A}\mathbb{B}^{\top}$?

2. Let $$\mathbb{A}=
\begin{bmatrix}
    1 &  3 & -2\\
    1 &  -1& 2
\end{bmatrix}
$$
and 
$$\mathbb{B}=
\begin{bmatrix}
    1 &  0\\
   -1 & 1\\
    2 & 1
\end{bmatrix}.
$$
Caculate $2\mathbb{A}\mathbb{B}$ by hand.
