---
title: "Quiz 1 generator, STATS 401 W18"
author: "Generates quizzes comparable to those in lab on 2/1 or 2/2"
output:
  html_document:
    theme: flatly
    toc: no
csl: ecology.csl
---

```{r preliminaries,echo=F,cache=F}
#ALL <- TRUE
ALL <- FALSE
#set.seed(99)
library(mat2tex)
```

\newcommand\mat[1]{\mathbb{#1}}
\newcommand\vect[1]{\mathbf{#1}}
\newcommand\transpose{{\scriptscriptstyle \mathrm{T}}}


**The quiz will test the skills covered in homeworks 1 to 4.  You will have 50 minutes allocated, though the quiz may take you much less time and you can leave lab once you are done.  The quiz will be closed book. Any electronic devices in your possession must be turned off and remain in a bag on the floor. Technical skills tested will include matrix multiplication and transposition,  inversion for 2x2, use of the sigma notation for sums. R coding will be tested by multiple choice questions. There will be questions on setting up equations for fitting a linear model, writing these equations in matrix form, and obtaining and interpreting the least squares fit to data.**

**This document generates different random quizzes each time it is compiled by Rmarkdown. The actual quiz will be a realization generated by this random process, or something very similar.**

---------------

### Matrix exercises

```{r randomMatrix,echo=F}
randomMatrix <- function(p,q,values=-4:4){
   matrix(sample(values,size=p*q,replace=TRUE),p,q)
}
```

```{r AB,echo=F}
A <- randomMatrix(3,2,values=-2:3)
B <- randomMatrix(2,2,values=-2:3)
```

M1. Evaluate $\mat{A}\mat{B}$ when 
```{r M1,echo=F,results='asis',cache=FALSE}
xx("\\mat{A}=",A,", \\quad \\mat{B} = ",B,mtype="bmatrix",digits=0)
```

M2. For $\mat{A}$ as above, write down $\mat{A}^\transpose$.

M3. For $\mat{B}$ as above, find $\mat{B}^{-1}$ if it exists.  If $\mat{B}^{-1}$ doesn't exist, explain how you know this.

------------

### Summation exercises

S1. A basic exercise.
```{r s1-all-list,include=F}
s1_files <- list.files(path="s1",full.names=T,pattern="*.Rmd$") 
```
```{r s1, child = sample(s1_files,size=1),eval=!ALL}
```
```{r s1-1, child = s1_files[1],eval=ALL}
```
```{r s1-2, child = s1_files[2],eval=ALL}
```
```{r s1-3, child = s1_files[3],eval=ALL}
```
```{r s1-4, child = s1_files[4],eval=ALL}
```
```{r s1-5, child = s1_files[5],eval=ALL}
```
```{r s1-6, child = s1_files[6],eval=ALL}
```

S2. An example involving sums of squares and products.
```{r s2-all-list,include=F}
s2_files <- list.files(path="s2",full.names=T,pattern="*.Rmd$")
```
```{r s2, child = sample(s2_files,size=1),eval=!ALL}
```
```{r s2-1, child = s2_files[1],eval=ALL}
```
```{r s2-2, child = s2_files[2],eval=ALL}
```
```{r s2-3, child = s2_files[3],eval=ALL}
```
```{r s2-4, child = s2_files[4],eval=ALL}
```
```{r s2-5, child = s2_files[5],eval=ALL}
```
```{r s2-6, child = s2_files[6],eval=ALL}
```
```{r s2-7, child = s2_files[7],eval=ALL}
```

### R exercises

R1. Using `rep()` and `matrix()`. 
```{r r1-all-list,cache=F,include=F}
r1_files <- list.files(path="r1",full.names=T,pattern="*.Rmd$")
```
```{r r1, child = sample(r1_files,size=1),eval=!ALL}
```
```{r r1-1, child = r1_files[1],eval=ALL}
```
```{r r1-2, child = r1_files[2],eval=ALL}
```
```{r r1-3, child = r1_files[3],eval=ALL}
```
```{r r1-4, child = r1_files[4],eval=ALL}
```

R2. Manipulating vectors and matrices in R.
```{r r2-all-list,cache=F,include=F}
r2_files <- list.files(path="r2",full.names=T,pattern="*.Rmd$")
```
```{r r2, child = sample(r2_files,size=1),eval=!ALL}
```
```{r r2-1, child = r2_files[1],eval=ALL}
```
```{r r2-2, child = r2_files[2],eval=ALL}
```
```{r r2-3, child = r2_files[3],eval=ALL}
```
```{r r2-4, child = r2_files[4],eval=ALL}
```

### Fitting a linear model by least squares 
```{r f-all-list,include=F}
f_files <- list.files(path="f",full.names=T,pattern="*.Rmd$")
```
```{r f, child = sample(f_files,size=1),eval=!ALL}
```
```{r f-1, child = f_files[1],eval=ALL}
```
```{r f-2, child = f_files[2],eval=ALL}
```
```{r f-3, child = f_files[3],eval=ALL}
```

------



Acknowledgements: Some questions are derived from https://genomicsclass.github.io/book. Some are derived from http://swirlstats.com/.


License: This material is provided under an [MIT license](https://ionides.github.io/401w18/LICENSE)<br>




------
