---
title: "Quiz1_Solutions"
author: "Naomi Giertych"
date: "2/17/2018"
output: pdf_document
---

# Matrix Exercises

M1. (2 points) Evaluate $\mathbb{AB}$ when

\begin{center}
$\mathbb{A} = \begin{bmatrix} -2 & 1 \\ 1 & 1 \\ -2 & 1 \end{bmatrix}$ $\mathbb{B} = \begin{bmatrix} -2 & -2 \\ 0 & 0 \end{bmatrix}$
\end{center}

$$
\mathbb{AB} = {\begin{bmatrix} 4 & 4 \\ -2 & -2 \\ 4 & 4 \end{bmatrix}}
$$

M2. (2 points) For $\mathbb{A}$ as above, write down $\mathbb{A}^T$

$$
\mathbb{A} = \begin{bmatrix} -2 & 1 & -2 \\ 1 & 1 & 1 \end{bmatrix}
$$

M3. (2 points) For $\mathbb{B}$ as above, find $\mathbb{B}^{-1}$ if it exists. If $\mathbb{B}^{-1}$ doesn't exist, explain how you know this.

(1 point) $\mathbb{B}^{-1}$ does not exist
(1 point) $\mathbb{B}^{-1}$ does not exist because $\frac 1{0-0}$ is undefined.

# Sumation exercises

S1. (3 points) A basic exercise.

Calculate $\sum_{i=k}^{k+4} (i+3)$, where $k$ is a whole number. Your answer should depend on k.

(2 points) $(k+3)+(k+4)+(k+5)+(k+6)+(k+7)$
(1 point) $5k + 25$

S2. (3 points) An example involving sums of squares and products.

Let $\mathbf{1} = (1,1, \dots, 1)$ and $\mathbf{x} = (x_1, x_2, \dots, x_n)$ be two vectors treated as $\mathbf{n \times 1}$ matrices. Use $\sum$ notation to evaluate teh matrix prodcut $\mathbf{1}^T\mathbf{x}$

$$
\sum_{i=1}^n x_i
$$

# R exercises

R1. Using rep() and matrix()

(3 points) Which of the following is the output of matrix(c(rep(0, times = 4), rep(1, times = 4)), ncol = 2)?

\begin{center}
(a) $\begin{bmatrix} 0 & 1 \\ 0 & 1 \\ 0 & 1 \\ 0 & 1 \end{bmatrix}$; (b) $\begin{bmatrix} 0 & 0 \\ 0 & 0 \\ 1 & 1 \\ 1 & 1 \end{bmatrix}$; (c) $\begin{bmatrix} 0 & 0 \\ 1 & 1 \\ 0 & 0 \\ 1 & 1 \end{bmatrix}$; (d) $\begin{bmatrix} 0 & 0 & 1 & 1 \\ 0 & 0 & 1 & 1 \end{bmatrix}$
\end{center}

A

R2. Manipulating vectors and matrices in R.

(3 points) Suppose x is a matrix in R. Which of the following is NOT equivalent to x?

(a). t(t(x))
(b). X %*% matrix(1,ncol(X))
(c). X*1
(d). X %*% diag(ncol(X))

B

# Fitting a linear model by least squares

F1. Fitting a linear model by least squares.

```{r}
library(faraway)
data("sat")
head(sat)
```

(2 points) Which of the following would produce the design matrix $\mathbb{X}$ for the model lm(sat ~ ratio + expend, data = sat).

(a) A <- matrix(rep(1, length(ratio)), ratio, expend)
(b) A <- matrix(1, ratio, expend)
(c) A <- cbind(rep(1, length(ratio)), ratio, expend)
(d) A <- cbind(1, ratio, expend)
(e) A <- cbind(ratio, expend)

C.

F2.(4 points) Consider our kicker data from homework 3.

```{r}
data_nfl <- read.csv("https://ionides.github.io/401w18/hw/hw03/FieldGoals2003to2006.csv",header = TRUE,skip=5)
head(data_nfl)
```

Recall that we built the model $y_i=mx_i+c_1z_{i,1}+c_2z_{i,2}+ \dots + c_{19}z_{i,19}+e_i$ where where $x_i$ is FGtM1 and $z_{i,1}$ takes the value 1 when row i of the data corresponds to kicker 1 (i.e., for i=1,2,3,4) and 0 otherwise. Write the design matrix of the model. (You do not need to include specific values for $x_i$.)

$$
\begin{bmatrix}
  x_1 & 1 & 0 & \dots & 0 \\
  x_2 & 1 & 0 & \dots & 0 \\
  x_3 & 1 & 0 & \dots & 0 \\
  x_4 & 1 & 0 & \dots & 0 \\
  x_5 & 0 & 1 & \dots & 0 \\
  x_6 & 0 & 1 & \dots & 0 \\
  x_7 & 0 & 1 & \dots & 0 \\
  x_8 & 0 & 1 & \dots & 0 \\
  \vdots & \vdots & \vdots & \dots & \vdots &\\
  x_n & 0 & 0 & \dots & 1 \\
\end{bmatrix}
$$

Points were allocated as follows:

(2 points) $x_i$ column

(1 point) for $z_{i,j}$'s correctly writen

(1 point) for $x_i$'s and $z_{i,j}$'s correctly labeled/described somewhere

(-1 point) for including 1 column for the intercept.