---
title: "Quiz 1 generator, STATS 401 W18"
author: "Generates quizzes comparable to those in lab on 2/1 or 2/2"
output:
  html_document:
    theme: flatly
    toc: yes
csl: ecology.csl
---

```{r preliminaries,echo=F,cache=F}
#set.seed(99)
library(mat2tex)
```

\newcommand\mat[1]{\mathbb{#1}}
\newcommand\transpose{{\scriptscriptstyle \mathrm{T}}}


**The quiz will test the skills covered in homework’s 1 to 4.  You will have 50 minutes allocated, though the quiz may take you much less time and you can leave lab once you are done.  The quiz will be closed book. Any electronic devices in your possession must be turned off and remain in a bag on the floor. Technical skills tested will include matrix multiplication and transposition,  inversion for 2x2, use of the sigma notation for sums. R coding will be tested by swirl-style multiple choice questions. There will be some questions on setting up equations for fitting a linear model, writing these equations in matrix form, obtaining and interpreting the least squares fit to data.**

**This document generates different random quizzes each time it is run (or, for each fixed value of the seed to the random number generator). The actual quiz will be a realization generated by this random process, or something very similar.**

---------------

### Matrix exercises

```{r randomMatrix,echo=F}
randomMatrix <- function(p,q,values=-4:4){
   matrix(sample(values,size=p*q,replace=TRUE),p,q)
}
```

```{r AB,echo=F}
A <- randomMatrix(3,2,values=-2:3)
B <- randomMatrix(2,2,values=-2:3)
```

1. Evaluate $\mat{A}\mat{B}$ when 
```{r M1,echo=F,results='asis',cache=FALSE}
xx("\\mat{A}=",A,", \\quad \\mat{B} = ",B,mtype="bmatrix",digits=0)
```

2. For $\mat{A}$ as above, write down $\mat{A}^\transpose$.

3. For $\mat{B}$ as above, find $\mat{B}^{-1}$ if it exists.  If $\mat{B}^{-1}$ doesn't exist, explain how you know this.

------------

### Summation exercises

1. A basic example.
```{r s1, child = sample(list.files(path="s1",full.names=T),size=1)}
```

2. An example involving sums of squares and products.
```{r s2, child = sample(list.files(path="s2",full.names=T),size=1)}
```

------

### R exercises

1. Using `rep()` and `matrix()`. 
```{r r1, child = sample(list.files(path="r1",full.names=T),size=1)}
```

2. Using `lm()`.
```{r r2, child = sample(list.files(path="r2",full.names=T),size=1)}
```

------

### Fitting a linear model by least squares 


------



License: This material is provided under an [MIT license](https://ionides.github.io/401w18/LICENSE)<br>


------
