---
title: "Lab Solutions"
author: "401 GSI team"
date: "3/8/2018 and 3/9/2018"
output:
  beamer_presentation:
  colortheme: dolphin
incremental: no
ioslides_presentation:
  incremental: no
slidy_presentation:
  incremental: no
fontsize: 10pt
---

```{r, warning = F}
# install.packages("MASS")
#Load library MASS
library(MASS)
#Load data crabs
data('crabs')

# add indicator variable to data for crab species
crabs$mu1 <- (crabs$sex == "M" )*1
crabs$mu2 <- (crabs$sex == "F")*1
```

# Q1) Constructing CIs in R
```{r}
# Obtain estimate of population mean
cl_crabs <- lm(CL~mu1+mu2-1, data = crabs)
summary(cl_crabs)
```

# Q2) Constructing a 99% confidence interval for the mean of Male crabs

$$\bar y \pm z_{\frac {\alpha}2}s.e(\bar y)$$
$$32.8510 \pm 2.56(0.7097)$$
$$(31.1974, 34.5046)$$

# Constructing a 99% confidence interval for the mean of Female crabs
$$\bar y \pm z_{\frac {\alpha}2}s.e(\bar y)$$
$$31.3600 \pm 2.56(0.7097)$$
$$(29.7064, 33.0136)$$


# Q3) Difference in Means

```{r}
crabs$mu3 <- 1
crabs$mu_diff <- crabs$mu1

bd_crabs2 <- lm(CL ~ mu3 + mu_diff - 1, data = crabs)
summary(bd_crabs2)
```

# Q4) Constructing a 95% confidence interval for the difference in means

- (note: I am using the normal approximation)

$$\bar d \pm z_{\frac {\alpha}2}s.e(\bar d)$$
$$ 1.4910 \pm 2.56(1.0037)$$
$$(-0.847621, 3.829621)$$

# Confidence Intervals for Future Values

- Motivating Question: What's the point of performing a regression?

# Confidence Intervals for Future Values

- A $100 (1-\alpha)$% **Confidence Interval** for a mean future value (or the regression line at) $\tilde y$ given values $\tilde x$:
    - $\hat y \pm t_{(\frac {\alpha}2, n-2)}s \sqrt{\frac 1n + \frac {(\tilde x - \bar x)^2}{\sum_{i=1}^n (x_i - \bar x)^2}}$

- A $100 (1-\alpha)$% **Prediction Interval** for a future value $\tilde y$ given values $\tilde x$:
     - $\hat y \pm t_{(\frac {\alpha}2, n-2)}s \sqrt{1 + \frac 1n + \frac {(\tilde x - \bar x)^2}{\sum_{i=1}^n (x_i - \bar x)^2}}$
    
- It is important to note that the confidence interval is narrower than the prediction interval
    - Why is this? (Hint: What do we know about means from 250?)
    
- Details can be found in sections 2.3 and 2.4 of Sheather

# Confidence Intervals for Future Values in R

Construct a 95% confidence interval and a 95% prediction interval for the crab's body depth given it is a blue crab with a carapace length of 45.

```{r}

crab_bd_reg <- lm(BD ~ sp + CL, data = crabs)
summary(crab_bd_reg)

```

# Confidence Intervals for Future Values in R

```{r}
x_star <- data.frame(sp = "B", CL = 45)

# confidence interval
predict(crab_bd_reg, x_star, interval = "confidence")

# prediction interval
predict(crab_bd_reg, x_star, interval = "prediction")

```
