---
title: "Stats 401 Lab 4"
author: "401 GSI team"

output:
  beamer_presentation:
  colortheme: dolphin
incremental: no
ioslides_presentation:
  incremental: no
slidy_presentation:
  incremental: no
---
# Outline

- Homework feedback
- Review on summation sign
- Sample quiz

# Homework feedback - mad_libs in swirl
```{r, eval = T, echo=T}
mad_libs <- function(...){
  # unpacking argument
  args <- list(...) 
  
  # assign them to variables
  place <- args[["place"]]
  adjective <- args[["adjective"]]
  noun <- args[["noun"]]
  
  paste("News from", place, "today where", adjective, 
        "students took to the streets in protest of the new", 
        noun, "being installed on campus.")
}
```

# Homework feedback - NA and 0

NA stands for not available, which means the value is missing. NA is not equal to 0. (Suppose x=0, then the value of 0 is known, hence not missing).
```{r, eval = T, echo=T}
X = cbind(c(NA,0),c(1,2));X
Y = matrix(1:4, nrow = 2);Y

```

# Homework feedback - NA and 0
```{r, eval = T, echo=T}
# NA plus/time other value will give us NA
X + Y
X %*% Y

# can use is.na() to check for NA
is.na(X)
```
# Homework feedback

Any other questions about the homework?

# Review - summation sign
  
$\sum_{i=1}^n x_i = x_1 + x_2 + ... + x_n$. You can always do this expansion if 
you are uncertain what to do.

Useful results to remember:

  - $\sum_{i=1}^n c x_i = c \sum_{i=1}^n x_i$
  - $\sum_{i=1}^n c = nc$
  - $\frac{d}{dc}\sum_{i=1}^n f(x_i,c) = \sum_{i=1}^n \frac{d}{dc} f(x_i,c)$
  
# Review - summation sign

Example : Calculate 
$\frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2$ where $\bar x = \frac{1}{n} \sum_{i=1}^n x_i$.

# Review - summation sign

Example : Calculate 
$\frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2$ where $\bar x = \frac{1}{n} \sum_{i=1}^n x_i$.

Solution:

\begin{equation}
\begin{aligned}
    \frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2 & = \frac{1}{n} \sum_{i=1}^n (x_i^2 - 2x_i \bar x + \bar x ^2)\\
    & = \frac{1}{n} \sum_{i=1}^n x_i^2 - \frac{2}{n}\bar x \sum_{i=1}^n x_i  + \frac{1}{n}\sum_{i=1}^n \bar x ^2)\\
    & = \frac{1}{n} \sum_{i=1}^n x_i^2 - 2 (\bar x)^2 + (\bar x)^2\\
    & = \frac{1}{n} \sum_{i=1}^n x_i^2 - (\bar x)^2
\end{aligned}
\end{equation}


# Review - summation sign

Example : Suppose $\mathbf{c} = (c_1,...,c_p)$ and $\mathbf{v} = (v_1,...,v_p)$, Use $\Sigma$ notation to evaluate the matrix product $\mathbf{c}^{\top}\mathbf{v}$

# Review - summation sign
Example : Suppose $\mathbf{c} = (c_1,...,c_p)$ and $\mathbf{v} = (v_1,...,v_p)$, Use $\Sigma$ notation to evaluate the matrix product $\mathbf{c}^{\top}\mathbf{v}$

Solution:
$\mathbf{c}^{\top}\mathbf{v} = \sum_{i=1}^p c_i v_i$

# Quiz outline
- Test the skills covered in HW 1 to 4
- 50 minutes; start at the beginning of next lab
- Closed book
- In today's lab we will do a sample quiz, which will be similar to the real quiz next week

# Sample quiz - Matrix exercises

Suppose we define $\mathbb{A}$ and $\mathbb{B}$ as follows,
```{r, eval = T, echo=F}
randomMatrix <- function(p,q,values=-4:4){
   matrix(sample(values,size=p*q,replace=TRUE),p,q)
}

set.seed(999)
A <- randomMatrix(3,2,values=-2:3)
B <- randomMatrix(2,2,values=-2:3)
```
```{r, eval = T, echo=T}
A
B
```

Calculate the matrices returned by following r command:

1. A %*% B

2. t(A)

3. solve(B)

# Sample quiz - Summation exercises

1. Calculate $\sum_{i=k}^{k+5} (i+3)$

\vspace{10mm}

2. Calculate $\frac{d}{dm}\sum_{i=1}^{n} (y_i -m x_i)^2$

# Sample quiz - R exercises

Which of the following code successfully construct the matrix 
$\mathbb{A} = \begin{bmatrix}1 & 1\\2 & 2\\3 & 3\end{bmatrix}$

A. A <- matrix(c(1,1,2,2,3,3) ,nrow=3)

B. A <- cbind(c(1,1),c(2,2),c(3,3))

\textbf{C. A <- t(matrix(c(1,1,2,2,3,3) ,nrow=2))}

D. A <- c(c(1:3),c(1:3))

# Sample quiz - Fitting a linear model by least squares

We look at the uswage data. Recall that
```{r, eval = T, echo=F}
library(faraway)
data("uswages")
```
```{r, eval = T, echo=T}
head(uswages, n=4)
```
We want to fit a linear model using wage as response, educ and exper as predictors.

# Sample quiz - Fitting a linear model by least squares
Which of the following code succesfully construct the matrix $\mathbb{X}$.

A. X <- matrix(uswages\$educ, uswages\$exper)

B. X <- matrix(rep(1,nrow(uswages)), uswages\$educ, uswages\$exper)

\textbf{C. X <- cbind(rep(1,nrow(uswages)), uswages\$educ, uswages\$exper)}

D. X <- cbind(uswages\$educ, uswages\$exper)

# Sample quiz - Fitting a linear model by least squares
If we want to fit the model using R function lm(), which of the following call is correct?

A. lm(wage ~ ., data = uswages)

B. lm(y ~ x, data = uswages)

C. lm(wage = educ + exper, data = uswages)

\textbf{D. lm(wage ~ educ + exper, data = uswages)}

\vspace{10mm}

Explain briefly how you would check whether your proposed solution in is correct in R. 