---
html_document:
  theme: flatly
title: "Sample Quiz 2"
output:
  html_document: default
  pdf_document: null
  toc: no
csl: ecology.csl
toc: no
word_document:
  toc: no
---
  
\newcommand\E{\mathrm{E}}
\newcommand\SE{\mathrm{SE}}
\newcommand\var{\mathrm{Var}}
\newcommand\cov{\mathrm{Cov}}
\newcommand\prob{\mathrm{P}}
\newcommand\mat[1]{\mathbb{#1}}
\renewcommand\vect[1]{\boldsymbol{\mathrm{#1}}}
\newcommand\transpose{{\scriptscriptstyle \mathrm{T}}}
\usepackage{amsmath}
  
### Formulas
  
* You are not allowed to bring any notes into the exam. 

* The following formulas will be provided. To use these formulas properly, you need to make appropriate definitions of the necessary quantities.

(1)$\quad\quad\mathbf{b}=\big(\mathbb{X}^\top \mathbb{X} \big)^{-1}\, \mathbb{X}^\top\mathbf{y}$

(2)$\quad\quad\var(\hat{\mathbf{\beta}})= \sigma^2 \big( \mathbb{X}^\top \mathbb{X} \big)^{-1}$

(3)$\quad\quad\var(\mathbb{A}\mathbf{Y})=\mathbb{A}\var(\mathbf{Y})\mathbb{A}^\top$

(4)$\quad\quad\var(X)=\E\big[ (X-\E[X])^2\big] = \E[X^2]-\big(\E[X]\big)^2$

(5)$\quad\quad\cov(X,Y)=\E\big[ \big(X-\E[X]\big)\big(Y-\E[Y]\big)\big] = \E[XY]-\E[X]\,\E[Y]$

(6)$\quad\quad$
The binomial ($n,p$) distribution has mean $np$ and variance $np(1-p)$.

\vspace{1cm}

From `?pnorm`:
```
pnorm(q, mean = 0, sd = 1)
qnorm(p, mean = 0, sd = 1)
q: vector of quantiles.
p: vector of probabilities.
```

### Q1. Calculating mean and variance, and making a normal approximation
Let $X_1,X_2,\dots,X_n$ be independent random variables follows Uniform(0,1) distribution. 
Find the mean and variance of $X_1$. Use this to find the mean and variance of $\bar{X}=\frac{1}{n}\sum_{i=1}^n X_i$. 
Now suppose $n=50$ and suppose that $\bar X$ is well approximated by a normal distribution. Find $\prob(0.45<\bar X<0.55)$. 
Write your answer as a call to `pnorm()`. Your call to pnorm may involve specifying any necessary numerical calculations that you can't work out without access to a computer or calculator.

### Q2. Comparing means using linear model
Let's consider the crabs data set we studied in lab. Recall that species(sp) is a factor with 2 levels Blue(B) and Orange(O). We want to study the difference of frontal lobe size(FL) of two species. 

```{r}
library(MASS)
data(crabs)
```
```{r}
head(crabs)
```

Consider the model $Y_i = \mu_1x_{Bi} + \mu_2x_{Oi} + \epsilon_i$,$i=1,...,200$. $Y_i$ is the FL of observation i. $x_{Bi}$ is 1 if $sp=B$ for obervation i and 0 otherwise. Similarly, $x_{Oi}$ is 1 if $sp=O$ for obervation i and 0 otherwise. $\epsilon_i$ are i.i.d with mean 0 and variance $\sigma^2$. This model is fitted in R using lm() function and the summary is provided as below.
```{r}
lm_crab <- lm(FL~sp-1, data=crabs)
summary(lm_crab)
```


(a). What do $\mu_1$ and $\mu_2$ measure in the above probability model?

(b). Build a 95% confidence interval for $\mu_1$ using normal approximation

(c). Recall in homework we know that the full estimated covariance matrix of $\mathbf{\hat{\mu}} = (\hat{\mu}_1,\hat{\mu}_2)$ can be found by  
```{r}
V <- summary(lm_crab)$cov.unscaled * summary(lm_crab)$s^2
V
```
Use V and information provided in summary(lm_crab) to build a 95% confidence interval for $\mu_1-\mu_2$

### Q3. Making and interpreting an F-test

Consider the birth weight data set we have seen in lab For this question, we will look at columns bwt(birth weight), lwt(mother's weight), age(mother's age) and race(mother's race, 1 for white, 2 for black and 3 for other).
```{r}
library(MASS)
data(birthwt)
```

```{r}
head(birthwt)
```
We want to study the relationship bwt and race using F test, while lwt and age are provided as confounding variables. Let the null hypothesis, $H_0$, be the probability model where bwt is modeled to depend linearly on lwt and age. Let $H_a$ be the probability model where $H_0$ is extended to include race as a factor, as fitted in R by
```{r}
lm_bw <- lm(bwt ~ lwt + age +factor(race), data = birthwt)
```
The results from summary(lm_bw) and anova(lm_bw) are as follows
```{r}
summary(lm_bw)
```
```{r}
anova(lm_bw)
```
(a). Write out the null and alternative hypotheses of the F test by completely specifying the probability models.

(b). Interpret the results in `anova(lm_gpa)`. Specifically, 
read the sample test statistic from R output, give the distribution of the model-generated test statistic under $H_0$, and explain how the resulting p-value is calculated and interpreted.

### Q4. Prediction using a linear model
We still look at the birthwt data in Q3. Consider the model fitted at $H_a$. Suppose we are interested in predicting the birthweight of a baby who has a 30-year-old white mother with weight 130. 

Specify a row matrix $\mathbf{x}^*$ so that $y^*=\mathbf{x}^*\mathbf{b}$ gives the least square predictor. Use summary(lm_bw) from Q3 to find $y^*$.

 