---
html_document:
  theme: flatly
title: "Sample Quiz 2"
output:
  html_document: default
  pdf_document: null
  toc: no
csl: ecology.csl
toc: no
word_document:
  toc: no
---

\newcommand\E{\mathrm{E}}
\newcommand\SE{\mathrm{SE}}
\newcommand\var{\mathrm{Var}}
\newcommand\cov{\mathrm{Cov}}
\newcommand\prob{\mathrm{P}}
\newcommand\mat[1]{\mathbb{#1}}
\renewcommand\vect[1]{\boldsymbol{\mathrm{#1}}}
\newcommand\transpose{{\scriptscriptstyle \mathrm{T}}}
\usepackage{amsmath}  
  
### Formulas
  
* You are not allowed to bring any notes into the exam. 

* The following formulas will be provided. To use these formulas properly, you need to make appropriate definitions of the necessary quantities.

(1)$\quad\quad\mathbf{b}=\big(\mathbb{X}^\top \mathbb{X} \big)^{-1}\, \mathbb{X}^\top\mathbf{y}$

(2)$\quad\quad\var(\hat{\mathbf{\beta}})= \sigma^2 \big( \mathbb{X}^\top \mathbb{X} \big)^{-1}$

(3)$\quad\quad\var(\mathbb{A}\mathbf{Y})=\mathbb{A}\var(\mathbf{Y})\mathbb{A}^\top$

(4)$\quad\quad\var(X)=\E\big[ (X-\E[X])^2\big] = \E[X^2]-\big(\E[X]\big)^2$

(5)$\quad\quad\cov(X,Y)=\E\big[ \big(X-\E[X]\big)\big(Y-\E[Y]\big)\big] = \E[XY]-\E[X]\,\E[Y]$

(6)$\quad\quad$
The binomial ($n,p$) distribution has mean $np$ and variance $np(1-p)$.

\vspace{1cm}

From `?pnorm`:
```
pnorm(q, mean = 0, sd = 1)
qnorm(p, mean = 0, sd = 1)
q: vector of quantiles.
p: vector of probabilities.
```

### Q1. Calculating mean and variance, and making a normal approximation
Let $X_1,X_2,\dots,X_n$ be independent random variables follows Uniform(0,1) distribution. 
Find the mean and variance of $X_1$. Use this to find the mean and variance of $\bar{X}=\frac{1}{n}\sum_{i=1}^n X_i$. 
Now suppose $n=30$ and suppose that $\bar X$ is well approximated by a normal distribution. Find $\prob(0.45<\bar X<0.55)$. 
Write your answer as a call to `pnorm()`. Your call to pnorm may involve specifying any necessary numerical calculations that you can't work out without access to a computer or calculator.

Solution:

$\mathbb{E}(X_1)=\int_0^1 x dx = 1/2$

$\mathbb{E}(X_1^2)=\int_0^1 x^2 dx = 1/3$

$Var(X_1)=\mathbb{E}(X_1^2)-(\mathbb{E}(X_1))^2=1/12$

Thus, $\mathbb{E}(\bar{X})=\mathbb{E}(\frac{1}{n}\sum_{i=1}^n X_i)=\frac{1}{n}\sum_{i=1}^n \mathbb{E}X_i=\mathbb{E}X_i=1/2$

$Var(\bar{X})=Var(\frac{1}{n}\sum_{i=1}^n X_i)=\frac{1}{n^2}Var(\sum_{i=1}^n X_i)=\frac{1}{n}Var(X_i)=\frac{1}{12n}$

$\prob(0.45<\bar X<0.55) = pnorm(0.55,mean=1/2,sd=sqrt(1/(12* 30)))-pnorm(0.45,mean=1/2,sd=sqrt(1/(12*30)))$

### Q2. Comparing means using linear model
Let's consider the crabs data set we studied in lab. Recall that species(sp) is a factor with 2 levels Blue(B) and Orange(O). We want to study the difference of frontal lobe size(FL) of two species. 

```{r}
library(MASS)
data(crabs)
```
```{r}
head(crabs)
```

Consider the model $Y_i = \mu_1x_{Bi} + \mu_2x_{Oi} + \epsilon_i$,$i=1,...,200$. $Y_i$ is the FL of observation i. $x_{Bi}$ is 1 if $sp=B$ for obervation i and 0 otherwise. Similarly, $x_{Oi}$ is 1 if $sp=O$ for obervation i and 0 otherwise. $\epsilon_i$ are i.i.d with mean 0 and variance $\sigma^2$. This model is fitted in R using lm() function and the summary is provided below.
```{r}
lm_crab <- lm(FL~sp-1, data=crabs)
summary(lm_crab)
```


(a). What does $\mu_1$ and $\mu_2$ measure in the above probability model?

Solution:

$\mu_1$ is the population mean frontal lobe size for blue crabs. $\mu_2$ is the population mean frontal lobe size for orange crabs.

(b). Build a 95% confidence interval for $\mu_1$ using normal approximation

Solution:

$(14.056-1.96*0.315,14.056+1.96*0.315)=(13.44,14.67)$

(c). Recall in homework we know that the full estimated covariance matrix of $\mathbf{\hat{\mu}} = (\hat{\mu}_1,\hat{\mu}_2)$ can be found by  
```{r}
V <- summary(lm_crab)$cov.unscaled * summary(lm_crab)$s^2
V
```
Use V and information provided in summary(lm_crab) to build a 95% confidence interval for $\mu_1-\mu_2$

Solution:

Let $a = (1,-1)^T$.

$Var(a^T\mathbf{\hat{\mu}})=a^TVar(\mathbf{\hat{\mu}})a=a^TVa=0.198$

$\hat{\mu}_1-\hat{\mu}_2=14.056-17.110=-3.054$ 

Thus we have the 95% C.I. $(-3.054-1.96*\sqrt{0.198},-3.054+1.96*\sqrt{0.198})=(-3.926,-2.182)$

### Q3. Making and interpreting an F-test

Consider the birth weight data set we have seen in lab For this question, we will look at columns bwt(birth weight), lwt(mother's weight), age(mother's age) and race(mother's race, 1 for white, 2 for black and 3 for other).
```{r}
library(MASS)
data(birthwt)
```

```{r}
head(birthwt)
```
We want to study the relationship bwt and race using F test, while lwt and age are provided as confounding variables. Let the null hypothesis, $H_0$, be the probability model where bwt is modeled to depend linearly on lwt and age. Let $H_a$ be the probability model where $H_0$ is extended to include race as a factor, as fitted in R by
```{r}
lm_bw <- lm(bwt ~ lwt + age +factor(race), data = birthwt)
```
The results from summary(lm_bw) and anova(lm_bw) are as follows
```{r}
summary(lm_bw)
```
```{r}
anova(lm_bw)
```
(a). Write out the null and alternative hypotheses of the F test by completely specifying the probability models.

Solution:

Under $H_0$, we have the model $Y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}+ \epsilon_i$,$i=1,...,189$. Where $Y_i$ as the birth weight for obervation i, $x_{i1}$ is the mother's weight while $x_{i2}$ is the age for observation i.  $\epsilon_i$ are i.i.d with mean 0 and variance $\sigma^2$.

Under $H_a$, we have $Y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}+\beta_3 x_{i3} + \beta_4 x_{i4}+ \epsilon_i$,$i=1,...,189$. $x_{i3}$ is 1 if ith mother is black and 0 otherwise. $x_{i4}$ is 1 if ith mother is neither white nor black and 0 otherwise. Other letters are defined same as the model under $H_0$.

For the F test, we are testing 

$H_0:\beta_3=\beta_4=0$ against $H_a:\beta_3\neq 0$ or $\beta_4\neq 0$


(b). Interpret the results in `anova(lm_gpa)`. Specifically, 
read the sample test statistic from R output, give the distribution of the model-generated test statistic under $H_0$, and explain how the resulting p-value is calculated and interpreted.

Solution:

The sample test statistics $f_{obs}=4.7799$. The model-generated test statistic F follows an F distribution of 2 numerator degree of freedom and 184 denominator degree of freedom under $H_0$. 

Then p-value is calculated by $\mathbb{P}(F>f_{obs})=0.009467<0.05$. Thus we will reject $H_0$ at 0.05 level. Namely, mothers' race has significant effect on birthweight of babies.

### Q4. Prediction using a linear model
We still look at the birthwt data in Q3. Consider the model fitted at $H_a$. Suppose we are interested in predicting the birthweight of a baby who has a 30-year-old white mother with weight 130. 

Specify a row matrix $\mathbf{x}^*$ so that $y^*=\mathbf{x}^*\mathbf{b}$ gives the least square predictor. Use summary(lm_bw) from Q3 to find $y^*$.

Solution:

$\mathbf{x}^*=(1,130,30,0,0)$.

Thus $y^*=2461.147+130*4.620+30*1.299=3100.717$