---
title: "Stats 401 Lab 1"
author: "Naomi Giertych"
date: "1/2/2018"
fontsize: 18pt
output: 
  beamer_presentation:
    incremental: false
    colortheme: "dolphin"
---

# Welcome!

- GSI: Naomi Giertych
- Email: ngierty@umich.edu
- Office Hours: TBD

# Lab Objectives

- Develop statistical computation skills (R) and
  mathematical skills
  
- Provide examples and excerises to help prepare
  you for homework, quiz, and exam questions

# Lab Expectations and Requirements

- (Printed) Homework is due at the start of class

- If you need to make up a lab:
  Please email me and make-up GSI in advance
  
# What will lab look like?

- Lab will consist of the following:

  1) Brief review of what was covered in lecture the previous       week
  2) A worked example (this may be combined with 1 if reasonable)
  3) An in lab excerice
  4) Lab ticket to complete and hand in (for attendance)
  
- If time permits, I'm willing to guide you in homework. However, I will not directly provide answers.

# Questions?

# About You!

>- Raise your hand if you've programmed in any language before
>- Raise your hand if you've opened R Studio before this class
>- Raise your hand if you know what a vector is

# Questions?

- Has everyone been able to install R and R Studio successfullly?

# Worked Example Using Swirl

```{r, eval = FALSE}

# Install Swirl
  install.packages('swirl')

# Call packages

  library(swirl)

# Use Swirl
  swirl()
  
# Navigate to Lesson 2 by selecting 1, 
  # then selecting 1 again.

```

#Worked Example Using A New Dataset

```{r, eval = TRUE}

  # Install and call required packages
    # note: I already have faraway
    # installed so I've commented it out
  # install.packages('faraway', dependencies = T)
  library(faraway)

  # Load Dataset
  data("uswages")
  
  # What if you want to know what all the variables in 
  # this dataset are?
  ?uswages
```

---

```{r, eval = TRUE}    
  # Let's find some specific values in the dataset
  
  # What's the wage of the 1500th observation?
  # Method 1:
  uswages[1500,1]
  
  # Method 2
  uswages$wage[1500]

```

---

```{r, eval = TRUE}  
  # What's the education of the 13th observation?
  # Method 1
  uswages[13,2]
  
  # Method 2
  uswages$educ[13]
  
  # What's the maximum wage found in the dataset?
  max(uswages$wage)
``` 

---

```{r, eval = TRUE} 

  # What's the difference between the wage in the 13th
  # observation and the max?
  
  # Method 1
  uswages$wage[13]

  # Difference
  7716.05 - 550.81
```

---

```{r, eval = TRUE} 
  # Method 2
  # How do I save a variable?
  wage_13 <- uswages$wage[13]
  max_wage <-  max(uswages$wage)
  
  # Difference
  max_wage - wage_13

```

# In Lab Activity

```{r, eval = FALSE}

  # Still using the faraway library

  # Use the SAT dataset
  data("sat"")

  # What was the highest total SAT score?

  # What was the lowest total SAT score?

  # Calculate the difference in scores.
  
```

# Key Takeaways

- Learning a new programming language can be frustrating and fun (sometimes at the same time)

- Commenting code is EXTREMELY important. Future you will thank you!

# Lab Ticket

- Using the salary dataset from the lab activity, find the following:

  1) What is the median salary?
  2) What is the mean salary?
  3) What do these numbers suggest about the distribution of salaries?
  
# Homework Hint

```{r, eval = TRUE}

  # Figuring out the median ratio of the last 15 states 
  # (alphabetically) in the SAT dataset.
  median(sat$ratio[36:50])

```