---
title: "Stats 401 Lab 3"
author: "401 GSI team"
date: "1/18/2018 and 1/19/2018"
output:
  beamer_presentation:
    colortheme: dolphin
    incremental: no
  ioslides_presentation:
    incremental: no
  slidy_presentation:
    incremental: no
---

```{r, set-options, echo = FALSE}
  options(width=50)
```

# Announcements

- Homework 2 is due today
  - Please remember to include "Sources" and "Please explain"
- First quiz is coming up! (Feb. 1st or 2nd)
- Homework 1 and 2 solutions will be posted Friday afternoon.
 
# Quick Review: Matrix Properties

- Addition
$$
\mathbb{A}+\mathbb{B}=
\begin{bmatrix}
    a_{11}+b_{11} &  a_{12}+b_{12}\\
    a_{21}+b_{21} &  a_{22}+b_{22}
\end{bmatrix}
$$

- Scalar multiplication
$$
s\mathbb{A}=
\begin{bmatrix}
    sa_{11} &  sa_{12}\\
    sa_{21} &  sa_{22}
\end{bmatrix}
$$

- Transpose
$$\mathbb{A}^{\top}=
\begin{bmatrix}
    a_{11} &  a_{21}\\
    a_{12} &  a_{22}
\end{bmatrix}
$$

- Matrix multiplication
$$
\mathbb{A}\mathbb{B}=
\begin{bmatrix}
    a_{11}b_{11}+a_{12}b_{21} &  a_{11}b_{12}+a_{12}b_{22}\\
    a_{21}b_{11}+a_{22}b_{21} &  a_{21}b_{12}+a_{22}b_{22}
\end{bmatrix}
$$

# Quick Review: Matrix Properties (cont.) 

- Inverse
$$\mathbb{A}^{-1}=\frac{1}{a_{11}a_{22}-a_{12}a_{21}}
\begin{bmatrix}
    a_{22} &  -a_{12}\\
    -a_{21} &  a_{11}
\end{bmatrix}
$$

# Quick Review: How to Input Matrices into R

- Take the following matrix:

$$
\mathbb{A}=
\begin{bmatrix}
    1 & 1 & 3\\
    1 & 2 & 1\\
    3 & 0 & 2
\end{bmatrix}
$$


```{r, eval = TRUE}

# R takes a column vector and 
# splits it into 3 different rows
matrix_by_col <- matrix(c(1,1,3,1,2,0,3,1,2),nrow=3)
matrix_by_row <- matrix(c(1,1,3,1,2,1,3,0,2),
                        byrow = TRUE, nrow=3)

```

# Quick Review: How to Input Matrices into R

- Don't forget to check that your output is correct!

```{r, eval = TRUE}

matrix_by_col
matrix_by_row

```

# Matrices in Action

- Solving a system of linear equations
- Recall the US wages dataset that we saw in Lab 01

```{r, eval = TRUE, echo = FALSE}

library(faraway)
data("uswages")
head(uswages)

```

# Matrices in Action (cont.)

- Write the sample version of the linear model for wages with all other variables as explanatory variables using vector notation.

Step 1) Notation:

Let ${x_{i1}}$, ${x_{i2}}$, \dots, ${x_{ip}}$ and ${y_i}$ be the values of predictor variable *j* and the wage of worker *i*, respectively.

Step 2) Recognize the hidden matrix:

Linear model for each individual is
$$
{y_1} = {b_0} + {b_1}{x_{11}} + {b_2}{x_{12}} + \dots + {b_p}{x_{1p}} + {e_1}
$$
$$
{y_2} = {b_0} + {b_1}{x_{21}} + {b_2}{x_{22}} + \dots + {b_p}{x_{2p}} +{e_2}
$$
$$
\vdots
$$
$$
{y_n} = {b_0} + {b_1}{x_{n1}} + {b_2}{x_{n2}} + \dots + {b_p}{x_{np}} +{e_n}
$$

- This looks like a system of linear equations that we can put into matrix form! 

# Matrices in Action (cont.)

Step 3) Define the matrices and vectors:

Let $$\mathbb{X}=
\begin{bmatrix}
    x_{11} &  x_{12} \\
    x_{21} &  x_{22} \\
    \vdots & \vdots \\
    x_{n1} & x_{n2}
\end{bmatrix}
$$
Let $\bold{y}$ = $({y_1}, {y_2}, ..., {y_n})$, $\bold{b}$ = $({b_1}, {b_2})$, and $\bold{e}$ = $({e_1}, {e_2}, ..., {e_n})$ be the vector of wages, predictor variables, and error terms. Then
$$\bold{y} = \mathbb{X}\bold{b} + \bold{e}$$

# Solving for the least squares estimate

- Recall the least squares estimate for the predictor variable coefficients is $$\bold{b} = (\mathbb{X^T}\mathbb{X})^{-1}\mathbb{X}\bold{y}$$

- Let's solve for $\bold{b}$ together.

# Solving for $\bold{b}$

- Construct $\mathbb{X}$ matrix corresponding to the linear equation from before.

```{r}

# Note that I've left out one of the 
# regional dummy varibles
# Question for students: Why did I need to do this?
attach(uswages)
X <- cbind(intercept=rep(1, length(educ)),
           educ, exper, race, 
           smsa, ne, mw, 
           so, pt)

head(X, n=3)

```

# Solving for $\bold{b}$

```{r}

solve(t(X) %*% X) %*% t(X) %*% uswages$wage

```

# Checking our result $\bold{b}$

```{r}

# Use the linear model function in R for 
# including all the variables
wage_lm <- lm(wage ~ ., data = uswages)
coef(wage_lm)

```

# In Lab Activity

- Using the library(faraway) and data("infmort"):

1) Construct the linear equation using vector and matrix notation.
2) Estimate the least squares estimate of $\bold{b}$ using the design matrix $\mathbb{X}$.
3) Check your estimate by using the lm function in R.

# Lab ticket

- Write your least squares estimate of the fitted values for the infant mortality data.