F1. Recall the dataset `uswages` containing ten variables on 2000 subjects from the 1988 Current Population Survey.
```{r, eval = T, echo=F}
library(faraway)
data("uswages")
```
```{r, eval = T, echo=T}
head(uswages, n=4)
```
Suppose we want to fit a linear model using wage as response, with years of education and years of experience as predictors.
Which of the following code succesfully construct the matrix $\mathbb{X}$ for a representation $\vect{y}=\mat{X}\vect{b}+\vect{e}$.

(a). `X <- matrix(uswages\$educ, uswages\$exper)`

(b). `X <- matrix(rep(1,nrow(uswages)), uswages\$educ, uswages\$exper)`

(c). `X <- cbind(rep(1,nrow(uswages)), uswages\$educ, uswages\$exper)`

(d). `X <- cbind(uswages\$educ, uswages\$exper)`

--------------

F2. If we want to fit the model using R function `lm()`, which of the following call is correct?

(a). `lm(wage ~ ., data = uswages)`

(b). `lm(y ~ x, data = uswages)`

(c). `lm(wage = educ + exper, data = uswages)`

Explain briefly how you would check whether your proposed solution is correct. 

---------------